---
title: "TSCI 5050 Fall 2019, Homework 01"
author: "Alex F. Bokov"
email: "bokov@uthscsa.edu"
date: "September 24, 2019"
output:
  html_document:
    smart: no
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
```

## Introduction. 

Please complete the following steps. Some of them only need to be done by
specific people and in those cases their names are highlighted. Some of the
items below are accompanied by questions to be answered within the comments of
your `data_characterization.R` scripts. **There is nothing to hand in-- I will
review your file on RStudio. Please let me know when you're ready for me to look
at it any time before the next class on October 1st.** Likewise, let me know if
you're stuck.

You should test your work before you move on to the next question. But your
script may give erors because it is not finished yet. One approach is to
temporarily comment out the lines that cause errors and use report-compiling as
a test. Another approach is to hit the `CTRL-SHIFT-HOME` 3-key combo or
`⌘-SHIFT-HOME` on the Mac (I think, correct me if I'm wrong). This will
highlight the script from the beginning to your current location in the script
(make sure the highlighted region does not end in the middle of a line of code).
Hit the `CTRL-ENTER` (or `⌘-ENTER`) 2-key combo to run just the highlighted part
of the script. By the time you reach the end of this homework, hopefully the
script will run from start to finish without errors.

## 1. EVERYONE: add some packages you will need.

Look for a line in your script that begins with `packages <- ` and change it to
look like this...

```{r packages}
packages <- c('GGally','tableone','pander');
```

## 2. <span style="color:teal">Noel</span>: create your renaming table.

_Not_ in your script but in your console, i.e. the bottom pane, do the following
command.
```{r varmap}
.maptemp <- data.frame(origname=dct0$column,varname=dct0$column,dispname=dct0$column);
write.csv(.maptemp,file='varmap.csv');
```

## 3. EVERYONE: review your renaming table

The columns should be: 

* `origname`: Matching __exactly__ the original names of the columns in your data. Do not
   change anything in this column.
* `varname`: For columns that you want to rename, put the new name here, otherwise leave them as they are. __Important: do not make any names with spaces or `-`. Use `_` or `.` instead__. 
* `dispname`: The display name, for plots and tables. This one may have spaces and other punctuation. 

As long as the above requirements are met, you can rename your variables in `varname` and `dispname` 
to anything that makes sense to you, in any language you want, etc. 

If all the variable names in `varmap.csv` are already acceptable to you, change one `varname` and one `dispname` anyway, for practice. __Reminder:__ if you need to edit `varmap.csv` you do that by downloading it from RStudio to your local computer: click on the checkbox next to it in the __Files__ tab on the lower right, click the __More__ menu, choose __Export__ and save it to your local computer. Open it in Excel, make your changes, save, click the __Upload__ button in the __Files__ tab and upload it back to RStudio, overwriting the original.

## 4. <span style="color:teal">Noel</span>: include the `varmap.csv` table into your script.

Find the lines in your code that say: 
``` {r varmap_location}
# characterization ----
set.caption('Data Dictionary');
``` 
and change them to look like this:
``` {r varmap_added}
map0 <- autoread ('varmap.csv');

# characterization ----
set.caption('Data Dictionary');
```

## 5. <span style="color:purple">Christine</span>, <span style="color:darkorange">Jennifer</span>, <span style="color:teal">Noel</span>: add the code that renames your variables
After `map0 <- autoread ('varmap.csv')` add the following lines:
```{r colrename}
dct0$column <- make.unique(unlist(submulti(dct0$column,map0,method = 'startsends')));
names(dat00) <- dct0$column;
```
<span style="color:purple">Note to Christine: you're very close, but try
comparing your version to the one above. If all else fails, replace your
variable renaming code with a copy-paste of the above.</span>

<span style="color:darkorange">Note to Jennifer: you have this, it's just commented out. I believe it's safe to
remove the `#`s now.</span>

## 6. <span style="color:teal">Noel</span>: set the random seed
After the above lines, add this:
```{r setseed}
set.seed(project_seed);
```

## 7. EVERYONE: create a developmental subsample.
After the line with `set.seed(project_seed)` add the following, including the
commented out text with questions and then answer those questions in your script
in the space provided. If you need more space, just insert more lines that start
with `#' `. I encourage you to run bits of code to see what each does and to
invoke help for the various commands, e.g. `?nrow`. Here's another tip: if you
have an object named `FOO` you can type `class(FOO)` to find out what kind of
object it is.
```{r dat01}
#' 
#' * Q: What does the command `nrow()` do?
#'     * A: 
#'          
#'          
#' * Q: What does the command `sample()` do? What are its first and second
#'      arguments for?
#'     * A: 
#'          
#'          
#' * Q: If `foo` were a data frame, what might the expression `foo[bar,baz]` do,
#'      what are the roles of `bar` and `baz` in that expression, and what would
#'      it mean if you left either of them out of the expression?
#'     * A: 
#'          
#'          
#' 
dat01 <- dat00[sample(nrow(dat00), nrow(dat00)/2),];
```

`dat01` is a smaller dataset randomly sampled from `dat00`. You will use it to 
develop your analysis while setting aside the rest of the data for prediction 
and hypothesis testing, to avoid bias and overfitting. Unless otherwise 
indicated, you will be using `dat01` for the rest of this script.

## 8. EVERYONE: select variables to put into your `predictorvars` and
`outcomevars` Each of them should have at least two variables that exist in your
dataset. If your data has a Subject ID variable, don't use that one. Your
variables are allowed to have missing values, but the fewer missing the better.
At least one of your `predictorvars` should be a discrete variable with 2-4
possible values besides `NA` (missing data). Your data dictionary will be
helfpul in choosing such variables. If you are using the demo dataset, you need
to choose ones different from the ones I chose.

Long variable names are tricky-- it's very easy to get one character wrong and
that's enough for it to stop matching your column names. Sometimes it's a 
character that's hard to see, like a trailing space (never leave trailing 
spaces inside quoted strings). So that's why we're using `map0` to rename your
variables to shorter versions.

## 9. EVERYONE: update the plot to use the subsample and your new variable names
If you already have a place in your code that starts with `ggpairs` replace that 
line with this:
```{r ggpairs}
ggpairs(dat01[,mainvars]);
```
Make sure that the above expression is active, not commented out.

If everything in previous parts works but this part (#9) fails, chances are
it's because you misspelled a variable when you were doing part #8 above. Or, 
maybe you are using the original variable names instead of the re-named ones 
resulting from part #5 above.

Problems with mis-named variables may produce errors that begin with `Error:
Can't find columns`.

The easy way to work around that is comment out the `ggpairs` line again, run the
rest of the report, and copy-paste the variable names you want from the data
dictionary it will print out right into your `predictorvars` and `outcomevars`
__remember to put quotes around each name, separate the quoted names with
commas, and don't permit any trailing spaces inside the quotes__. Or if you have
run `source(data_characterization.R)` lately, instead of compiling the report
you could print out the column names right in your console with `dct0$column`
and then copy them from the console to `predictorvars` and `outcomevars` in your
script. Did I mention, __remember to put quotes around each name, separate the
quoted names with commas, and don't permit any trailing spaces inside the
quotes__? 

Once you added the correct variable nameds to `predictorvars` and `outcomevars`
you can remove the `#` from the `ggpairs` line and try compiling your report 
again.

## 10. EVERYONE: Create a cohort table.
`CreateTableOne` creates a summary of your patients (or samples) of the sort 
people like to put at the beginnings of their research papers, hence the name.

```{r tableone}
#' * Q: Which function 'owns' the argument `caption`? What value does that 
#'      argument pass to that function?
#'     * A: 
#'          
#'          
#' * Q: Which function 'owns' the argument `printToggle`? What value does that 
#'      argument pass to that function?
#'     * A: 
#'          
#'          
#' * Q: Which function 'owns' the argument `vars`? We can see that the value
#'      this argument passes comes from the variable `mainvars`... so what is
#'      the actual value that ends up getting passed to the function?
#'     * A: 
#'          
#'          
#' * Q: What is the _very first_ argument of `print()` in the expression below?
#'      (copy-paste only that argument into your answer without including 
#'      anything extra)
#'     * A: 
#'          
#'          
pander(print(CreateTableOne(vars = mainvars, data = dat01, includeNA = TRUE)
             , printToggle=FALSE)
       , caption='Cohort Characterization');
```


Good luck!

